---
title: "Death trends"
output:
  html_document:
    theme: paper
    mathjax: null
    toc: true
    self_contained: false
    code_download: true
    includes:
      in_header: header.html
---

```{css, echo=F}
pre.r > code {white-space: pre;}
```

# Code

Imports.

```{r setup, message=F}
library(CausalImpact)
library(gghighlight)
library(scales)
library(zoo)
```

Main logic.

```{r logic}
go <- function(deaths) {
  deaths <- na.trim(deaths)
  data <- deaths
  data$Cumulative <- ave(data,
                         format(time(data), "%Y"),
                         FUN=cumsum)
  names(data) <- c("Monthly", "Cumulative")
  data <- fortify.zoo(data, melt=T)
  data$Year <- format(data$Index, "%Y")
  data$Month <- as.numeric(format(data$Index, "%m"))
  data$Index <- NULL

  p <- function() {
    ggplot(data) +
      geom_line(aes(x=Month, y=Value, group=Year, color=Year)) +
      gghighlight(Year %in% c("2020", "2019"),
                  use_group_by=F,
                  calculate_per_facet=T) +
      facet_wrap(~Series, dir="v", scales="free_y") +
      xlab("") + ylab("Deaths") +
      scale_y_continuous(labels=scales::label_number_si()) +
      scale_x_discrete(limits=month.abb, guide=guide_axis(check.overlap=T)) +
      theme_bw(base_size=15)
  }

  print(suppressMessages(p()))

  data <- deaths
  if (any(is.na(data))) {
    message("Interpolated missing data.")
    data <- na.approx(data, na.rm=T)
  }

  tryCatch({
    pre.period <- c(start(data), as.yearmon("Dec 2019"))
    post.period <- c(as.yearmon("Jan 2020"), end(data))
    model.args <- list(niter=1000, nseasons=12)

    impact <- CausalImpact(data,
                           pre.period,
                           post.period,
                           model.args=model.args)

    cat(paste(impact$report, "\n\n"))

    print(plot(impact) +
            scale_y_continuous(labels=scales::label_number_si()) +
            ylab("Deaths"))

    impact$summary["Cumulative",]
  }, error=message)
}

total <- function(deaths) {
  deaths <- na.approx(deaths)
  data <- zoo(cbind(rowSums(deaths)), time(deaths))
  names(data) <- "V1"
  go(data)
}

by.region <- function(deaths) {
  excess <- data.frame()
  for (region in names(deaths)) {
    nice.region <- gsub("[.]", " ", region)
    cat(paste("\n\n####", nice.region, "\n\n"))
    data <- subset(deaths, select=region)
    row <- go(data)
    if (!is.null(row)) {
      row.names(row) <- c(nice.region)
      excess <- rbind(excess, row)
    }
  }
  excess
}

plot.excess <- function(excess, level="####") {
  excess$Country <- row.names(excess)
  excess <- excess[order(-excess$RelEffect),]

  cat(paste("\n\n", level, " Relative\n\n", sep=""))
  p <- ggplot(excess, aes(x=RelEffect, y=Country)) +
    geom_col() +
    geom_errorbar(aes(xmin=RelEffect.lower, xmax=RelEffect.upper)) +
    gghighlight(p < 0.05 & !(RelEffect.lower <= 0 & 0 <= RelEffect.upper)) +
    scale_x_continuous(labels=scales::label_percent()) +
    xlab("") + ylab("") +
    theme_bw()

  print(p)

  excess <- excess[order(-excess$AbsEffect),]

  cat(paste("\n\n", level, " Absolute\n\n", sep=""))
  p <- ggplot(excess, aes(x=AbsEffect, y=Country)) +
    geom_col() +
    geom_errorbar(aes(xmin=AbsEffect.lower, xmax=AbsEffect.upper)) +
    gghighlight(p < 0.05 & !(AbsEffect.lower <= 0 & 0 <= AbsEffect.upper)) +
    scale_x_continuous(labels=scales::label_number_si()) +
    xlab("") + ylab("") +
    theme_bw()

  print(p)
}
```

Download the [datasets](https://docs.google.com/spreadsheets/d/1eSaKBxUfWB1payf2OjEZt5x83CaeK_RIuq8u1heji9E/edit?usp=sharing).

```{r data}
get.sheet.url <- function(gid) {
  paste("https://docs.google.com/spreadsheets/d/e/2PACX-1vQfgIZSsUrFzbmVK0hjbTjvY2TGI75FRkrU8rGKpLebZnfFFBCvFyI1zPgUE3uivCOmbOKOYJWyYUiQ/pub?gid=",
        gid,
        "&single=true&output=csv",
        sep="")
}

au.deaths <- read.csv.zoo(get.sheet.url("1434566235"), FUN=as.yearmon)
nz.deaths <- read.csv.zoo(get.sheet.url("894713646"), FUN=as.yearmon)
us.deaths <- read.csv.zoo(get.sheet.url("1095978846"), FUN=as.yearmon)
uk.deaths <- read.csv.zoo(get.sheet.url("0"), FUN=as.yearmon)

# Save the output of `total` from every country so we can display excess deaths
# statistics for each country at the end.
countries <- data.frame()
```

# Content

## Australia {.tabset}

Sources

-   [Provisional Mortality Statistics](https://www.abs.gov.au/statistics/health/causes-death/provisional-mortality-statistics/latest-release#data-download)

    -   Provisional mortality statistics weekly dashboard, Jan-Oct 2020;
    -   Doctor certified deaths by week of occurrence, 2015-19.

### National

```{r au.country, results="asis"}
countries <- rbind(countries, Australia=total(au.deaths))
```

### Regional {.tabset .tabset-pills}

```{r au.region, results="asis"}
au.region <- by.region(au.deaths)
```

### Excess deaths {.tabset .tabset-pills}

```{r au.excess, results="asis"}
plot.excess(au.region)
```

## New Zealand {.tabset}

Source

-   [COVID-19 data portal](https://www.stats.govt.nz/experimental/covid-19-data-portal)

### National

```{r nz.country, results="asis"}
countries <- rbind(countries, `New Zealand`=total(nz.deaths))
```

### Regional {.tabset .tabset-pills}

```{r nz.region, results="asis"}
nz.region <- by.region(nz.deaths)
```

### Excess deaths {.tabset .tabset-pills}

```{r nz.excess, results="asis"}
plot.excess(nz.region)
```

## UK {.tabset}

Source

-   [Deaths registered monthly in England and Wales](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/monthlyfiguresondeathsregisteredbyareaofusualresidence)

### National

```{r uk.country, results="asis"}
countries <- rbind(countries, `United Kingdom`=total(uk.deaths))
```

### Regional {.tabset .tabset-pills}

```{r uk.region, results="asis"}
uk.region <- by.region(uk.deaths)
```

### Excess deaths {.tabset .tabset-pills}

```{r uk.excess, results="asis"}
plot.excess(uk.region)
```

## US {.tabset}

Source

-   [Underlying Cause of Death, 1999-2019
Request](https://wonder.cdc.gov/ucd-icd10.html). Centers for Disease Control and
Prevention, National Center for Health Statistics. Underlying Cause of Death
1999-2019 on CDC WONDER Online Database, released in 2020. Data are from the
Multiple Cause of Death Files, 1999-2019, as compiled from data provided by the
57 vital statistics jurisdictions through the Vital Statistics Cooperative
Program. Accessed at <http://wonder.cdc.gov/ucd-icd10.html> on Jan 3, 2021
6:49:16 AM

-   [Weekly Counts of Deaths by State and Select Causes, 2020-2021](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/muzy-jte6)

Note: I couldn't find pre-2020 data for Puerto Rico, so it's excluded from impact analysis.

### National

```{r us.country, results="asis"}
countries <- rbind(countries,
                   `United States`=total(subset(us.deaths, select=-Puerto.Rico)))
```

### Regional {.tabset .tabset-pills}

```{r us.region, results="asis"}
us.region <- by.region(us.deaths)
```

### Excess deaths {.tabset .tabset-pills}

```{r us.excess, results="asis", fig.height=6}
plot.excess(us.region)
```

# Global excess deaths {.tabset}

```{r country.excess, results="asis"}
plot.excess(countries, level="##")
```
